---
title: "Competing risks"
author: "Graeme L. Hickey, Pete Philipson, Ruwanthi Kolamunnage-Dona, Paula Williamson"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Competing risks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Description

As of version 2.0, `joineR` fits the joint model proposed by Williamson et al. 
(2008) for joint models of longitudinal data and competing risks. In Williamson 
et al. (2008), the longitudinal data submodel remained as per that analyzed in 
Henderson et al. (2000), namely

$$Y_{ij} = x_{ij} \beta + a_{ij}^\top U_i + Z_{ij},$$

where $Y_{ij}$ is the $j$-th repeated measurement on subject $i$, $U_i$ is a 
latent vector that follows a zero-mean multivariate normal distribution, 
$x_{ij}$ and $a_{ij}$ are vectors of explanatory variables that may be 
time-constant or time-varying, and the $Z_{ij}$ are mutually independent errors,
$Z_{ij} \sim {\rm N}(0,\tau^2)$.

For the competing risks data, a cause-specific hazards submodel is proposed, 
namely

$$\lambda_{ig}(t) = \lambda_{0g}(t) \exp(w_{ij} \alpha_g + b_{ij}^\top 
V_{ig}),$$

where $w_{ij}$ and $b_{ij}$ are vectors of explanatory variables that may be 
time-constant or time-varying, $V_i$ is another latent vector that follows a
zero-mean multivariate normal distribution, and $\lambda_{0g}(t)$ is the baseline hazard function for cause $g = 1, 2$.

In `joineR`, when competing risks data are analysed, $a_{ij}^\top  = (1, 
t_{ij})$, whilst the random effects for the repeated measurement and 
time-to-event sub-models are proportional, hence $b_{ij}^\top V_{ig} = \gamma_g 
a_{ij}^\top U_i$. We also note that currently on 2 failure types are be 
modelled, which must be coded as 1 and 2 (with 0 representing censoring).


# Example

We will perform the analysis from Williamson et al. (2008) to the SANAD (Standard
And New Antiepileptic Drugs) study data -- the largest trial of AEDs to date 
(Marson et al., 2007).

## Background

The time to withdrawal of a randomized drug or addition of another, has been
recommended by the International League Against Epilepsy to be one of the
primary endpoints for clinical trials of anti-epileptic drugs (AEDs) (Commission
on Antiepileptic Drugs, 1998). Patients may decide to switch to an alternative
AED because of inadequate seizure control (ISC) or to withdraw from a treatment
because of unacceptable adverse effects (UAE). Overall analysis of treatment
failures may miss differential effects of AEDs on the reasons for withdrawal,
which may differ in terms of their relative importance for patients (Williamson
et al., 2007).

The primary SANAD analysis of patients with partial epilepsy concluded that the 
newer drug lamotrigine (LTG) was preferable in terms of treatment failure to 
carbamazepine (CBZ), which had been the standard for many years (Williamson et
al., 2006). LTG was significantly better in terms of withdrawal for UAE and not
significantly worse in terms of withdrawal for poor seizure control. Subsequent
correspondence after the trial results were published indicated that some
readers were concerned that differential titration rates may have been to the
disadvantage of CBZ (Cross et al., 2007). An AED that is titrated more quickly
may bring benefits in terms of seizure control but be more likely to cause
adverse effects. This criticism has also been leveled at previous AED trials
(O’Donoghue & Sander, 1995).

## Data 

SANAD was an unblinded randomized trial recruiting patients with epilepsy for
whom CBZ was considered to be standard treatment and they were randomized to
CBZ, gabapentin, LTG, oxcarbazepine or topiramate (Marson et al., 2007). Time to
treatment failure was a primary outcome with the analysis of competing risks,
i.e. treatment failure due to either ISC or UAE, an important secondary
objective. Although LTG appeared to be superior to CBZ in the main analysis,
concerns were raised that the results were biased in favor of the newer drug.
CBZ was considered to have been titrated more quickly bringing benefits in terms
of seizure control but being more likely to cause adverse effects. The data set
analysed here, comparing CBZ and LTG, includes 605 patients. During the trial,
94 patients withdrew from the randomized drug due to UAE while 120 withdrew due
to ISC. In order to compare the two AEDs after adjusting for titration rate,
dose calibration was first undertaken by standardizing the dose of both drugs
relative to the midpoint of the maintenance dose range for each particular drug.
The maintenance dose recommended in the SANAD trial was independently deemed to
be reasonable (Faught, 2007) and the approach to calibration considered
sensible. These calibrated doses are taken to be the longitudinal measurements
in the competing-risks joint model.

The data is available in the `joineR` package by running

```{r epileptic_data}
library(joineR)
data(epileptic)
head(epileptic)
```

In the analysis, Williamson et al. (2008) considered an interaction term between
time and treatment in the longitudinal data submodel. Therefore, we will
manually code this term using

```{r interaction_data}
epileptic$interaction <- with(epileptic, time * (treat == "LTG"))
```

A `jointdata` object is first constructed using the `jointdata()` function. Note
that there are several status columns: `with.status` is a composite event status
(treatment failure or censored), `with.status.uae` is a binary event status for
UAE (with failure due to ISC treated as censoring), `with.status.isc` is a
binary event status for ISC (with failure due to UAE treated as censoring), and
`with.status2` is a vector taking values 0, 1, and 2, with 0 denoting censoring
and 1 and 2 denoting dropout due to ISC and UAE, respectively.

```{r jointdata, fig.width=7, fig.height=4}
longitudinal <- epileptic[, c(1:3, 13)]
survival <- UniqueVariables(epileptic, c(4, 6), "id")
baseline <- UniqueVariables(epileptic, "treat", "id")
data <- jointdata(longitudinal = longitudinal,
                  survival = survival,
                  baseline = baseline,
                  id.col = "id",
                  time.col = "time")

summary(data)
jointplot(data, Y.col = "dose", Cens.col = "with.status2")
```

## Model

To fit the joint model, we use the usual `joint()` function as

```{r jointmodel, cache=TRUE}
fit2 <- joint(data = data,
              long.formula = dose ~ time + treat + interaction,
              surv.formula = Surv(with.time, with.status2) ~ treat,
              longsep = FALSE, survsep = FALSE,
              gpt = 3)

summary(fit2)
```

To estimate standard errors, we use the `jointSE()` function as follows.
However, we do not run the code in this vignette due to computational time
required.

```{r jointmodel_ses, cache=TRUE, eval=FALSE}
fit2.se <- jointSE(fit2, n.boot = 100)
fit2.se
```

## Conclusion

Significant $\gamma$ estimates in the model suggests that calibrated dose is
associated with time to treatment failure for both particular causes. It is
observed that these estimates have opposite signs for the two failure reasons.
This was a situation anticipated previously in this application, since side
effects are more likely at higher doses, whereas seizure control is poorer at
low doses (Williamson et al., 2006). Thus, patients on higher doses may be more
likely to be withdrawn from treatment due to poor seizure control since the
reason they are on higher doses is that they are having continued seizures.
Patients on higher doses may also be less likely to be withdrawn from treatment
due to UAE since the reason a required dose increase is possible is because no
such events have occurred

The results from all models fitted suggest that if LTG is titrated at the same
rate as CBZ, the beneficial effect of LTG on UAE would still be evident and the
two drugs still appear to provide similar seizure control. Thus, the conclusion
from the original simpler analysis of the cause-specific hazards is found to be
robust to the variation in titration between the drugs in the trial.


# Acknowledgements

The `joineR` package was funded by the UK Medical Research Council (MRC) under a
grant with Principal Investigator (PI) Prof. Paula Williamson; Co-Investigators
(Co-Is) Prof. Peter J. Diggle and Prof. Robin Henderson; and Research Associates
Dr Ruwanthi Kolamunnage-Dona, Dr Peter Philipson, and Dr Ines Sousa (Grant
numbers G0400615). Updates to it were made under a separate MRC grant with PI Dr
Ruwanthi Kolamunnage-Dona, Co-Is Dr Peter Philipson and Dr Andrea Jorgensen, and
Research Associate Dr Graeme L. Hickey (Grant number MR/M013227/1).


# References

Williamson PR, Kolamunnage-Dona R, Philipson P, Marson AG. Joint modelling of longitudinal and competing risks data. *Statistics in Medicine*, 2008; **27**: 6426-6438.

4. Henderson R, Diggle PJ, Dobson A. Joint modelling of longitudinal measurements
and event time data. *Biostatistics*, 2000; **1**: 465-480

1. Marson AG, Al-Kharusi AM, Alwaidh M, Appleton R, Baker GA, Chadwick DW, Cramp
C, Cockerell OC, Cooper PN, Doughty J, Eaton B, Gamble C, Goulding PJ, Howell
SJL, Hughes A, Jackson M, Jacoby A, Kellett M, Lawson GR, Leach JP, Nicolaides P,
Roberts R, Shackley P, Shen J, Smith DF, Smith PEM, Tudur Smith C, Vanoli A,
Williamson PR on behalf of the SANAD Study group. Carbamazepine, gabapentin,
lamotrigine, oxcarbazepine or topiramate for partial epilepsy: results from the
SANAD trial. *The Lancet*, 2007; **369**: 1000–1015.

2. Commission on Antiepileptic Drugs. Considerations on designing clinical
trials to evaluate the place of new antiepileptic drugs in the treatment of
newly diagnosed and chronic patients with epilepsy. *Epilepsia*, 1998; **39**:
799–803.

3. Williamson PR, Smith CT, Josemir WS, Marson AG. Importance of competing risks
in the analysis of anti-epileptic drug failure. *Trials*, 2007; **8**: 12.

4. Williamson PR, Kolamnunnage-Dona R, Tudur Smith C. The influence of competing
risks setting on the choice of hypothesis test for treatment effect.
*Biostatistics*, 2006; **8**: 689–694.

5. Cross H, Ferrie C, Lascelles K, Livingston J, Mewasingh L. Old versus new
antiepileptic drugs: the SANAD study. *The Lancet*, 2007; **370**: 314.

6. O’Donoghue MF, Sander JWAS. Lamotrigine versus carbamazepine in epilepsy.
*The Lancet*, 1995; **345**: 1300.

7. Faught E. Epilepsy drugs: getting it right the first time. *Lancet Neurology*,
2007; **6**: 476–478.
